---
title: "Introduction to rKolada"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rKolada}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`rKolada` is an R package for *downloading*, *inspecting* and *processing* data from [Kolada](https://kolada.se/), a Key Performance Indicator database for Swedish municipalities and regions. This vignette provides an overview of the methods included in the `rKolada` package and the design principles of the package API. To learn more about the specifics of functions and to see a full list of the functions included, please see `??rKolada`. For a quick introduction to the package see the vignette [A quick start guide to rKolada](quickstart-rkolada.html).

NOTE: All metadata and data labels in Kolada are written in Swedish only.

The design of `rKolada` functions is inspired by, and are supported by, the design and functionality provided by several packages in the `tidyverse` family. It is thus recommended that you install the `tidyverse` package before installing rKolada:

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("rKolada")
```

## Kolada, a Key Performance Indicator database for Swedish municipalities and regions

The Swedish Municipalities and Regions Database [Kolada](https://kolada.se/) is a openly accessible, comprehensive database containing over 4,000 Key Performance Indicators (KPIs) for a vast number of aspects of municipal and regional organisations, politics and economic life. The `rKolada` R package provides an interface to R users to directly download, explore, and simplify metadata and data from Kolada.

To get started with Kolada you might want to visit its homepage (Swedish-only) or read through the [REST API documentation on Github](https://github.com/Hypergene/kolada). However, you can also use the `rKolada` package to explore data without prior knowledge of the database.

```{r setup}
library("rKolada")
```

### The data model

Data in Kolada are stored along three basic _dimensions_:

- A KPI ID
- A point in time (year)
- A municipality/region/ ID

When downloading data, the user needs to specify search parameters for at least two of these three basic dimensions. (The Kolada API documentation also specifies a fourth basic dimension: _gender_. However, data for all genders is always automatically downloaded when available.) The parameters can be a single, atomic value or a vector of values.

Also, the Kolada database proves useful groupings of municipalities and KPIs that can be used for further exploration, or to create unweighted averages. Lastly, some KPIs are also available for _Organizational units_ (OUs) within municipalities, e.g. a school, an administrative subdivision or an elderly home.


### Downloading data

If the user already has knowledge of the IDs of the KPIs and/or municipalities they want to download, this can be done using the function `get_values()`. For instance, if you want to download all values for the KPI `N00945` ("Tillfälliga föräldrapenningdagar (VAB) som tas ut av män, andel av antal dagar (%)") for Sweden's three most populous cities; Stockholm (id `"0180"`), Gothenburg (Swedish: _Göteborg_; `"1480"`) and Malmö (`"1280"`):

```{r}
n00945 <- get_values(
  kpi = "N00945",
  municipality = "G123833",
  period = 1970:2020
)

n00945
```

In many cases, however, you will not know in advance exactly what KPIs to be looking for, or you might not know the IDs of Sweden's municipalities.

### Downloading metadata: `get` functions

Kolada has five different kinds of metadata entities Each one of these can be downloaded by using `rKolada`'s `get` functions. Each function returns a `tibble` with all available data for the specified metadata entity:

- KPIs: `get_kpi()`
- KPI groups: `get_kpi_groups()`
- Municipalities: `get_municipality()`
- Municipality groups: `get_municipality_groups()`
- Organizational Unit: : `get_ou()`

Each function returns a `tibble` with all available data for the specified metadata entity.

```{r}
# Download all KPI metadata as a tibble (kpi_df)
kpi_df <- get_kpi()

kpi_df
```

All `get` functions are thin wrappers around the more general function `get_metadata()`. If you are familiar with the terminology used in the Kolada API for accessing metadata you might want to use this function instead.

### Exploring metadata

For each metadata type mentioned in the previous sections, `rKolada` offers several convenience functions to help exploring and narrowing down metadata tables. (If you are familiar with `dplyr` semantics, most of these functions are basically wrappers around `dplyr`/`tidyr` code.)

Since each `get()` function above returns a table for the selected entity, a metadata table can be one of five different types. All metadata convenience functions are prefixed to reflect which kind of metadata table they operate on: `kpi`, `kpi_grp`, `municipality`, `municipality_grp`, and `ou`.

All metadata convenience functions have been designed with functional piping in mind, so their first argument is always a metadata tibble. Most of them also return a tibble of the same type

The most important family of metadata convenience functions is the `search()` family. Much like `dplyr::filter()` they can be used to search for one or several search terms in the entire table or in a subset of named columns:

```{r}
# Search for KPIs with the term "BRP" in their description or title
kpi_filter <- kpi_df %>% kpi_search("BRP", column = c("description", "title"))
```


